Genetic Dilution erodes productivity: Exploring farmers' low adoption levels of improved maize in Ethiopia

### 2023-03-20 paper.tex
Updates from Overleaf

% 	\affil[1]{\small \emph{The World Bank}}
% 	\affil[2]{\small \emph{FAO}}
% 	\affil[3]{\small \emph{Assistant Research Professor at the Center for Data Science for Enterprise and Society, Cornell University}}
\author{%
 Aleksandr Michuda \footnote{Center for Data Science for Enterprise and Society, Cornell University}%
 \and Cristina Chiarella\footnote{UC Louvain}%
 \and Oscar Barriga-Cabanillas\footnote{The World Bank}%
  \and Juan Sebastian Correa \footnote{FAO}%
  }
%   Assistant Research Professor at the 
	\date{\today}

    \clearpage\maketitle
	\thispagestyle{empty}
	\vspace*{-2em}
    \begin{center}\begin{abstract}
			\noindent  
			Despite large investments in  developing new  maize varieties, evidence shows that few farmers consistently adopt this new technology over time. We use an innovative methodology that studies the adoption trajectories of farmers over a three-round panel data set to explain low adoption rates with heterogeneous returns to adoption. By relying on self-reported use of improved seeds, we find a puzzling answer for why farmers do not use new seed varieties: adoption leads to seemingly small and negative returns. Heterogeneity analysis across different levels of comparative advantage show that farmers who adopt don't actually have a comparative advantage in adopting improved seed, nor do they exhibit high returns to adopting. To reconcile this result with the expected higher yields for improved seeds, we exploit unique information on maize DNA-fingerprinting collected over the same areas in 2018. We find that the negative returns to adoption mask heterogeneity in the actual quality of the genetic material of the seeds farmers categorized as improved varieties in the self-reported data. Positive returns to adoption are found for those farmers that use higher-purity germplasm, drought-tolerant maize, and newly released varieties. When we account for misclassification in our heterogeneity analysis, we find that the puzzle is resolved. Our findings point to the wide dispersion of older and genetically diluted varieties, for which poorer farmers may be paying a premium. The implications of our findings speak to the need for policies to better target context and geography, expand accessibility of improved seeds, and make higher yields varieties more inclusive.

@@

%%% Why it is important
Food security is a major challenge in many East African countries, and Ethiopia is no exception, having historically struggled to provide an adequate and reliable food supply \citep{Ramakrishna2002-hv, Jaleta2018-oj}. Since the drought of 1984, Ethiopia has become increasingly reliant on maize, and as of today, maize constitutes the most widespread crop in the country.\footnote{FAO statistics show the total area cultivated with Maize in 2019 is 25\% larger than the next most widespread crop (sorghum), and 27\% and 139\% larger than the next two following crops, wheat and barley, respectively.} As Figure \ref{fig:maize_yields} shows, there is an upward trend in the total area cultivated with maize since 1990, with average yields almost doubling during this period. This increase in maize yields has coincided with a dramatic decrease in undernourishment of the population which has fallen from 47\% in 1990 to 14\% in 2018 (The World Bank, 2021). But despite these positive gains in total area planted and average yields, food insecurity still affects more than half the population of Ethiopia.\footnote{The country has also recently seen increases in food insecurity because of COVID-19, the conflict in Ethiopia’s Tigray region and the desert locust outbreak.} Moreover, a large productivity gap persists between the African continent and the rest of the developing world. The persistence of this gap is often attributed to different adoption rates of green revolution technologies, which include improved seeds. Evidence has shown that the adoption of improved seed leads to increases in agricultural yields \citep{Carter2014-fm}, food security \citep{Shiferaw2014-op}, and poverty reduction \citep{Minten2008-tj}.\footnote{Nevertheless, critics of green revolution technologies have also claimed that these technological packages have not been as profitable for many farmers in Africa as promised (with the exception of Ethiopia), given the high costs, and their heavy dependence on subsidies \citep{Wise20}.}

\begin{figure}[htp]
    \centering
    \caption{Maize: Total area harvested and average yields} \label{fig:maize_yields}
    \includegraphics[width=.6\textwidth]{results/figures/Maize_yields.pdf}
    \vspace*{-1em}
    \begin{table}[H]
        \centering
        \begin{tabular}{p{0.6\textwidth}} 
            \begin{tablenotes}[flushleft]
                  \small
                  \item Source: Authors' calculations using FAOSTAT data 
                  \item Note: Yields are measured as the harvested production per ha for the area under cultivation
            \end{tablenotes}
        \end{tabular}
    \end{table}       
\end{figure}

@@

Despite evidence and numerous technological breakthroughs in maize germplasm with increased yields and drought tolerance, only 13\% of maize farmers in Ethiopia have adopted improved seeds consistently between 2011 and 2015. A large percentage has adopted new varieties inconsistently, and the largest share of farmers (62\%) have not used improved seeds at all during the same period. Several answers have been proposed to explain such low adoption of improved varieties. These include lack of liquidity and other economic constraints \citep{Carter2014-fm}, growing conditions such as the type of soil, input use, and farmer characteristics \citep{Munshi2004-og}, imperfections in credit markets \citep{Croppenstedt2003-pq}, ambiguously defined property rights \citep{Place2000-el}, lack of commitment devices \cite{Duflo2009-iv}, high transportation costs that increase the cost of agricultural inputs \citep{Byerlee2013-qk}, and improved seed supply constraints \citep{Bird2020-nt}. Additionally, the net returns to adoption of improved seeds are often lower than expected in the long term because a large percentage of farmers unknowingly use diluted improved seed varieties, or ones that are over 20 years old. With such old seed, there is evidence that yield improvement and disease resistance wanes over time\citep{Abate2015-rj}. The variety of these constraints and their dispersion across subpopulations and locations speaks to considerable heterogeneous returns to adoption.

@@

In this paper, we revisit the mechanism proposed by \citep{Suri2011-oi}, where a lack of adoption of improved seed varieties is the result of farmers' heterogeneous returns to the new technology. Ignoring such heterogeneity hides the true impact of adoption, as farmers with different characteristics and skill levels observe different returns from adoption despite the technology providing high average returns. The heterogeneous returns to adoption are modeled as a farmer's unobservable comparative advantage using a Correlated Random Coefficient (CRC) model. In the original paper, \citep{Suri2011-oi} finds that farmers’ lack of adoption is driven by insufficient net benefits to the technology stemming from poor infrastructure that increases the cost of access. \footnote{Similar work in Ethiopia on improved chickpea varieties suggests that the economic returns to adoption plays a major role in the decision to adopt. \citep{Michler2018-wk}.}

%% Expand a bit on the methods
We build on this literature by investigating improved maize seed adoption in Ethiopia. In particular, we implement two complementary estimation strategies to address both how heterogeneous returns drive the decision to use improved seed varieties, and the extent that the misclassification of varieties and genetic dilution affect the decision to adopt them by lowering their observed returns. First, we implement a Group Random Coefficients estimation, which builds on \cite{Suri2011-oi} in \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} using a General Method of Moments (GMM) Estimator that allows us to test \textbf{ the assumptions made by \cite{Suri2011-oi}}, as well as provide more economically interpretable results. The advantage of our estimation strategy is that we do not rely on identification from a valid instrument. %\textbf{to assess XXXXXX}. %\footnote{Our estimation strategy is similar to \cite{Chamberlain1984-uk} \textbf{XXXXXX}.}

For the first set of results, our empirical strategy relies on the first three survey rounds of the Ethiopian Socioeconomic Survey (ESS). The data were collected as a panel, which cover 2011, 2013, and 2015. For each round, farmers self-reported on the seed variety they used (improved or traditional). By using the adoption trajectories over three rounds, we find the returns to adoption are in most cases, not statistically significant, and only significant with wave 3 adoption. This posits a puzzle that motivates our next estimation strategy.

We expand our results using the fourth round of the ESS (conducted in 2018), which provides
a unique opportunity to understand why improved seed varieties seem not to increase returns
when adopted. The ESS4 collected DNA fingerprints of harvested cropcuts in selected plots and compared its genetic material with a reference library of available varieties in Ethiopia. This data provides critical information not only about the seed types but also about their purity, sources, and their year of release. We exploit this uniquely rich information to compare different definitions of improved seeds. Specifically, we compare returns of verifiably improved seeds against self-reported use of new, second generation or recycled varieties, comparing different purity thresholds (or degrees of no-contamination with other varieties), year of release, types of varieties, and their sources. Because the latest survey round is not a continuation of the first three rounds, our second estimation strategy implements an Endogenous Switching Regression (ESR) approach, a standard estimation strategy to study the question of adoption with cross-sectional data (\citealt{Marenya2020-kb}, \citealt{Shiferaw2014-op}), given its capacity to account for unobserved heterogeneity in adoption. The results of the ESR show negative effects of adopting improved seeds, as per farmer self-reports. But the comparison of different improved seed definitions using the DNA fingerprinting shows that when the definition of improved seed is based on greater purity of the germplasm (95\% of purity or more), the returns to adoption for those that used such improved varieties becomes positive and significant. We obtain similar results when evaluating the specific effects of adopting drought-tolerant maize varieties, CGIAR or Exotic varieties, or more recently released varieties.

@@

Finally, to integrate our results together, we use the misclassification rates from the 4th round of the data and re-estimate the GRC results to be robust to misclassification. This method takes inspiration from \cite{michuda2021three} to account for the fact that the self-reports from the first three rounds of the ESS data may be misclassified and attenuated. We find that once we account for misclassification, heterogeneity in adoptions becomes very similar to other contexts; higher comparative advantage is associated with lower returns pointing to significant barriers to adoption for those that would benefit the most from it. Incorporating misclassification rates from the fourth round of the ESS into the GRC estimation resolves the puzzle posed by the GRC results.

@@

We contribute to this literature by exploring heterogeneous comparative advantages to adoption in Ethiopia between 2011 to 2015. %, controlling for important drivers of adoption such as rainfall (found to be the main driver of heterogeneous effects for adopting agronomic packages \citealt{Marenya2020-kb}, \citealt{Katengeza2019-af}), labor and input use. 
Our findings provide an alternative hypothesis to the puzzle of low maize adoption, namely that the wide dispersion of older generations and recycled improved seeds, may be resulting in lower yields and altering farmers' input use. The null and negative effects that we see in the GRC results are actually stemming from misclassification in self-reported adoption status. This paper is the first to take seriously the bias that this can create when estimating the treatment effect of adoption, both in a homogeneous and heterogeneous setting. This also implies a methodological discovery when investigating hybrid seed adoption: the absence of a true control group to use for analysis. With more time and more genetic dilution, it becomes more likely that even those that do report hybrid use are still likely using some level of hybrid.

@@

\subsection{The Three Round Panel}

@@

Table \ref{tbl:summary} shows a set of summary statistics for the sample. The mean parcel size for a household is about 0.15 Ha, with a maximum size of a little less than 1.5 Ha. Harvest labor is disproportionately skewed to family participation with around 20 days of harvest labor being conducted by family members and only about 2 days for hired labor. This is indicative of either \textbf{malfunctioning} labor markets, income constraints or high supervision costs. The distances to some key areas are also highlighted in Table \ref{tbl:summary}. On average, households are around 13 kilometers away from an asphalt road and almost 60 kilometers away from the nearest market. This makes both the purchase of improved seed and the marketing of yields subject to transportation costs and a difficult enterprise. Fertilizer costs are on average 8,350 Birr ($\thicksim$178 USD), with a large amount of the sample not using fertilizer at all. This is particularly important in the context of improved maize seed use as these technologies are intensive in external inputs, such as chemical fertilizers and irrigation, to be effective. Only 5\% of the sample irrigates their crops making households susceptible to weather shocks.

@@

A seed is considered “improved” in Ethiopia if it was tested by breeders and evaluated to be superior to existing varieties  (\citep{MoA13}; \citep{kosmowski2020shining}). In Ethiopia, the government controls most of the seed system, with local agricultural offices or extension agents assessing the demands of different varieties. The MoA decides the seed quantities to be produced, and actors such as regional seed enterprises, seed companies, seed unions, are involved in the production of certified seeds, which are later distributed by seed producer cooperatives (SPCs).\footnote{They account for 37\% of the total distribution in the country (\citep{kosmowski2020shining}).} An improved seed can be obtained directly from such distribution centers (which would distribute a first generation seed), however it is often the case that seeds cultivated from second or later generations from the originally bred seed are also exchanged between farmers. In the survey, farmers are asked to report whether the variety they used was a first-generation improved seed, second generation, recycled, or a traditional variety. When buying seeds, farmers might be able to identify the variety they are looking for, but it is unlikely that they would be aware of the source.

Fifty four varieties have been released in Ethiopia since 1990.\footnote{Out of which seven were released between 1990 to 1999, 29 between 2000 to 2009, and 18 between 2010 to 2019} Thirty four such varieties contain International Maize and Wheat Improvement Center (CIMMYT) germplasm, out of which, 10 Drought-tolerant maize varieties (DTMZ) and 8 Quality protein maize (QPM) were released (most of these between 2010 to 2019). Besides the CIMMYT-related germplasm, ten are open-pollinated varieties (OPVs) and 25 are hybrid (\citep{kosmowski2020shining}).

The fourth wave of the ESS introduced DNA fingerprinting of barley, maize, and sorghum for the first time. The collection of cropcuts for crop yield estimation facilitated the process for DNA fingerprinting. For the cropcuts, a 4 square meter quadrant was randomly selected in each plot, and total production harvested from the quadrant was weighed (both when freshly harvested and when dried). The DNA material was extracted from samples from the dried cropcuts, and were then matched with the genetic material of the maize DNA reference library for Ethiopia, obtained from a CIMMYT and Ethiopian Institute of Agricultural Research (EIAR) project. The collection of DNA samples focused on eight regions. In each enumeration area (EA) of these regions, a random sample of plots was selected, with a maximum of 10 fields for each crop. As a result, 505 plots were selected for DNA fingerprinting, corresponding to 447 households.

@@

We find that 38\% of households self-report adopting a “new” improved variety (SR2 in Figure \ref{fig:adoption_r4}), and 43\% report adopting a “new”, “second generation” or “recycled” improved variety (SR1). Considering a household as adopter if the household planted an improved variety in at least one of their plots, leads to classifying all households where a DNA sample was obtained as users of improved seed variety, a result that holds when the purity threshold of 70\% (DNA 70) is used. It is only when using purity thresholds of 90 and 95\% (DNA 90, DNA 95 in Figure \ref{fig:adoption_r4}, respectively), that adoption rates of improved varieties change to 91 and 54\% of sampled households, respectively.\footnote{We also find that 62\% of households adopted a hybrid variety, compared to 38\% that adopted an open-pollinated variety, and 15\% of households adopted a drought-tolerant maize variety, while 63\% of households that adopted an improved variety use a CGIAR-derived germplasm, and 79\% use an exotic germplasm.} Finally, 64\% of households adopted a variety that was released in the year 2000 or after, and 24\% adopted a variety that was adopted in the year 2010 or after. These ratios are in line with prior findings (\cite{Zeng15}).

@@

We find regional variations in adoption levels. In the regions of Oromia, Amhara, Tigray, and SNNPR, around 39\% of households report adopting an improved varierty . In the regions of Amhara, Benishangul-Gumuz, Oromia, SNNPR, and Tigray, adoption rates of improved maize varieties were found for 27\% of households (\cite{Jaleta2018-oj}). In Oromia, SNNPR, and Benishangul-Gumuz adoption of improved varieties per farmers’ self-reports was found for 84\% of maize growers in 2011, and 88\% of maize growers in 2013 (\cite{Yirga17}).

@@

Since ESS1-ESS3 do not have DNA fingerprinting, we are left with only self-reports of hybrid use. We use these variables in a Group Random Coefficients (GRC) approach to investigate the heterogeneous effects of adoption. The advantage of the GRC approach is that we do not rely on identification from a valid instrument, but on modeling the heterogeneity directly through a projection of adoption returns on their observed ``trajectory'' of adoption. This allows us to directly capture two effects: heterogeneous effects in the form of the so-called ``comparative advantage" of adoption, and, under the assumption of linearity in comparative advantage, a description of the overall economy's sorting of farmers based on their comparative advantage. The first is important because it is able to disentangle the effect of a general household's agricultural ability (which can be thought of as a fixed effect) and their adoption-specific potential. The second parameter can tell us whether the economy sort farmers in a way that is commensurate with providing those that would benefit the most from adoption actually adopting, or whether there are barriers interfering with that. See \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} for more information on the assumptions involved for the GRC approach.

@@

The endogenous switching regression holds certain advantages over other methodologies used for identifying unbiased estimates with cross-sectional observational data such as propensity score matching or other instrumental variable approaches (\cite{Shiferaw2014-op}) and has been used to study similar questions in the past (\citealt{falco2011does}, \citealt{kabunga2012yield}). \par
This approach models the effect of improved seed adoption on yields in two stages. First, the decision to adopt an improved seed is determined by a series of observed and unobserved variables. Equation \ref{eq:switch_latent} shows the definition for a latent variable $A^*$ that reflects the benefit from adopting an improved seed relative to planting a traditional one. As shown in the equation, household $i$ will decide to adopt ($A_i$=1) if these perceived benefits of adoption are positive.

@@

The results of the regression are shown in Table \ref{tbl:unres}. The results for self-reported and cropcut yields are largely similar. There is some discrepancy in the self-reported yields in the first wave, and specifically with the $(1,0,0)$ trajectory. The return to adoption is negative for cropcuts, but positive for self-reported yields. The returns to adoption in many cases, are not statistically significant, or they are negative. Only in the case of the trajectory $(0,0,1)$: those that adopted in the second and third waves, is there a large, statistically significant effect.

>>>>

% 	\affil[1]{\small \emph{The World Bank}}
% 	\affil[2]{\small \emph{FAO}}
% 	\affil[3]{\small \emph{Assistant Research Professor at the Center for Data Science for Enterprise and Society, Cornell University}}
\author{%
 Aleksandr Michuda \footnote{Center for Data Science for Enterprise and Society, Cornell University}%
 \and Cristina Chiarella\footnote{UCLouvain}%
 \and Oscar Barriga-Cabanillas\footnote{The World Bank}%
  \and Juan Sebastian Correa \footnote{FAO}%
  }
%   Assistant Research Professor at the 
	\date{\today}

    \clearpage\maketitle
	\thispagestyle{empty}
	\vspace*{-2em}
    \begin{center}\begin{abstract}
			\noindent  
			Despite large investments in  developing new  maize varieties, evidence shows that few farmers consistently adopt this new technology over time. We study the adoption trajectories of farmers over a three-round panel data set to explain low adoption rates with heterogeneous returns to adoption. By relying on self-reported use of improved seeds, we find a puzzling answer for why farmers do not use new seed varieties: adoption leads to seemingly small and negative returns. Heterogeneity analysis across different levels of comparative advantage show that farmers who adopt do not have a comparative advantage in adopting improved seed, nor do they exhibit high returns to adopting. To reconcile this result with the expected higher yields for improved seeds, we exploit unique information on maize DNA-fingerprinting collected over the same areas in 2018. We find that the negative returns to adoption mask heterogeneity in the actual quality of the genetic material of the seeds farmers categorized as improved varieties in the self-reported data. Positive returns to adoption are found for those farmers that use higher-purity germplasm, drought-tolerant maize, and newly released varieties. When we account for misclassification in our heterogeneity analysis, we find that the puzzle is resolved. Our findings point to the wide dispersion of older and genetically diluted varieties, for which poorer farmers may be paying a premium. The implications of our findings speak to the need for policies to better target context and geography, expand accessibility of improved seeds, and make higher yields varieties more inclusive.

@@

%%% Why it is important
Food security is a major challenge in many East African countries, and Ethiopia is no exception, having historically struggled to provide an adequate and reliable food supply \citep{Ramakrishna2002-hv, Jaleta2018-oj}. Since the drought of 1984, Ethiopia has become increasingly reliant on maize, and as of today, maize constitutes the most widespread crop in the country.\footnote{FAO statistics show the total area cultivated with Maize in 2019 is 25\% larger than the next most widespread crop (sorghum), and 27\% and 139\% larger than the next two following crops, wheat and barley, respectively.} There was an upward trend in the total area cultivated with maize since 1990, with average yields almost doubling during this period. \footnote{Improved security conditions and maize productivity contributed with an important decrease in undernourishment which fell from 47\% in 1990 to 14\% in 2018 (The World Bank, 2021). Despite these positive gains in total area planted and average yields, food insecurity still affects more than half the population of Ethiopia.The country has also recently seen increases in food insecurity because of COVID-19, the conflict in Ethiopia’s Tigray region and the desert locust outbreak.} Despite progress a large productivity gap persists between the African continent and the rest of the developing world. From 2016 to the present, cereal crop yields in Sub-Saharan Africa have consistently been around 70-80\% lower than Europe or North America \citep{worldbankyield}. The persistence of this gap is often attributed to different adoption rates of green revolution technologies, which include improved seeds. Evidence has shown that the adoption of improved seed leads to increases in agricultural yields \citep{Carter2014-fm}, food security \citep{Shiferaw2014-op}, and poverty reduction \citep{Minten2008-tj}.\footnote{Nevertheless, critics of green revolution technologies have also claimed that these technological packages have not been as profitable for many farmers in Africa as promised. Even the most successful of these countries, Ethiopia, is not on track to meet the goals of the Alliance for a Green Revolution for Africa (AGRA). This is due to high costs, and heavy dependence on subsidies \citep{Wise20}.}

% \begin{figure}[htp]
%     \centering
%     \caption{Maize: Total area harvested and average yields} \label{fig:maize_yields}
%     \includegraphics[width=.6\textwidth]{results/figures/Maize_yields.pdf}
%     \vspace*{-1em}
%     \begin{table}[H]
%         \centering
%         \begin{tabular}{p{0.6\textwidth}} 
%             \begin{tablenotes}[flushleft]
%                   \small
%                   \item Source: Authors' calculations using FAOSTAT data 
%                   \item Note: Yields are measured as the harvested production per ha for the area under cultivation
%             \end{tablenotes}
%         \end{tabular}
%     \end{table}       
% \end{figure}

@@

Despite evidence and numerous technological breakthroughs in maize germplasm with increased yields and drought tolerance, only 13\% of maize farmers in Ethiopia have adopted improved seeds consistently between 2011 and 2015. A large percentage has adopted new varieties inconsistently, and the largest share of farmers (62\%) have not used improved seeds at all during the same period. Several answers have been proposed to explain such low adoption of improved varieties. These include lack of liquidity and other economic constraints \citep{Carter2014-fm}, growing conditions such as the type of soil, input use, and farmer characteristics \citep{Munshi2004-og}, imperfections in credit markets \citep{Croppenstedt2003-pq}, ambiguously defined property rights \citep{Place2000-el}, lack of commitment devices \citep{Duflo2009-iv}, high transportation costs that increase the cost of agricultural inputs \citep{Byerlee2013-qk}, and improved seed supply constraints \citep{Bird2020-nt}. Additionally, the net returns to adoption of improved seeds are often lower than expected in the long term because a large percentage of farmers may  use diluted improved seed varieties, or ones that have been recycled over many generations. This is a concern as there is evidence that yield improvement and disease resistance wanes over time \citep{Abate2015-rj}. The variety of these constraints and their dispersion across subpopulations and locations speaks to considerable heterogeneous returns to adoption.

@@

In the absence of a natural experiment or randomized intervention, it is difficult to properly estimate the effect of adoption as there is potential selection of certain farmers into adoption improved varieties. This selection bias is more pervasive when the goal is to explore heterogeneous effects of adoption. Alternatively, one can make assumptions about the structure of the underlying error distribution to properly identify returns to adoption. In this paper, we revisit the latter mechanism proposed by \cite{Suri2011-oi}, where a lack of adoption of improved seed varieties is the result of farmers' heterogeneous returns to the new technology. Ignoring such heterogeneity hides the true impact of adoption, as farmers with different characteristics and skill levels observe different returns from adoption despite the technology providing high average returns. The heterogeneous returns to adoption are modeled as a farmer's unobservable comparative advantage. In the original paper, \cite{Suri2011-oi} finds that farmers’ lack of adoption is driven by insufficient net benefits to the technology stemming from poor infrastructure that increases the cost of access. Similar work in Ethiopia on improved chickpea varieties suggests that the economic returns to adoption plays a major role in the decision to adopt. \citep{Michler2018-wk}.

%% Expand a bit on the methods
% We build on this literature by investigating improved maize seed adoption in Ethiopia. In particular, we implement two complementary estimation strategies to address both how heterogeneous returns drive the decision to use improved seed varieties, and the extent that the misclassification of varieties and genetic dilution affect the decision to adopt them by lowering their observed returns. First, we implement a Group Random Coefficients estimation, which builds on \cite{Suri2011-oi} and \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} using a General Method of Moments (GMM) estimator that allows us to test \textbf{ the assumptions made by \cite{Suri2011-oi}}, as well as provide more economically interpretable results. The advantage of our estimation strategy is that we do not rely on identification from a valid instrument. %\textbf{to assess XXXXXX}. %\footnote{Our estimation strategy is similar to \cite{Chamberlain1984-uk} \textbf{XXXXXX}.}

Our results are divided into two parts. We start by using the first three survey rounds of the Ethiopian Socioeconomic Survey (ESS), which is a panel  covering  2011, 2013, and 2015. For each round, farmers self-reported on the seed variety they used (improved or traditional). By using the adoption trajectories over three rounds, we find the returns to adoption are in most cases, not statistically significant, and only significant for adoption in the third wave of the survey. This posits a puzzle that motivates our next estimation strategy.

Our second set of results use the fourth round of the ESS (conducted in 2018), which collected DNA fingerprints of harvested cropcuts in selected plots and compared its genetic material with a reference library of available varieties in Ethiopia, hence, providing a unique opportunity to understand why improved seed varieties seem not to increase returns when adopted. These data provides critical information not only about the seed types but also about their purity, sources, and their year of release. We exploit this uniquely rich information to compare returns of verifiable improved seed varieties against self-reported use of new, second generation or recycled varieties. For this, we leverage different purity thresholds (or degrees of no-contamination with other varieties), year of release, types of varieties, and their sources. Because the latest survey round is not a continuation of the first three rounds, our second estimation strategy implements an Endogenous Switching Regression (ESR) approach, a standard estimation strategy to study the question of adoption with cross-sectional data (\citealt{Marenya2020-kb}, \citealt{Shiferaw2014-op}), given its capacity to account for unobserved heterogeneity in adoption. The results of the ESR show negative effects of adopting improved seeds, as per farmer self-reports. But the comparison of different improved seed definitions using the DNA fingerprinting shows that when the definition of improved seed is based on greater purity of the germplasm (95\% of purity or more), the returns to adoption for those that used such improved varieties becomes positive and significant. We obtain similar results when evaluating the specific effects of adopting drought-tolerant maize varieties, CGIAR or Exotic varieties, or more recently released varieties.

@@

Finally, to integrate our results together, we use the misclassification rates from the 4th round of the data and re-estimate the GRC results to be robust to misclassification. This method takes inspiration from \cite{michuda2021three} to account for the fact that the self-reports from the first three rounds of the ESS data may be misclassified and attenuated. We find that once we account for misclassification, heterogeneity in adoptions becomes very similar to other contexts: higher comparative advantage is associated with lower returns, pointing to significant barriers to adoption for those that would benefit the most from it. Incorporating misclassification rates from the fourth round of the ESS into the GRC estimation resolves the puzzle posed by the GRC results.

@@

We contribute to this literature by exploring heterogeneous comparative advantages to adoption in Ethiopia between 2011 to 2015. %, controlling for important drivers of adoption such as rainfall (found to be the main driver of heterogeneous effects for adopting agronomic packages \citealt{Marenya2020-kb}, \citealt{Katengeza2019-af}), labor and input use. 
Our findings provide an alternative hypothesis to the puzzle of low maize adoption, namely that the wide dispersion of older generations and recycled improved seeds may be resulting in lower yields and altering farmers' input use. The null and negative effects that we see in the GRC results are actually stemming from misclassification in self-reported adoption status. This paper is the first to take seriously the bias that this can create when estimating the treatment effect of adoption, both in a homogeneous and heterogeneous setting. This also implies a methodological discovery when investigating hybrid seed adoption: the absence of a true control group to use for analysis. With more time and more genetic dilution, it becomes more likely that even those that do report hybrid use are still likely using some level of hybrid.

@@

\subsection{The Three Round Panel and the Fourth Wave}

@@

Table \ref{tbl:summary} shows a set of summary statistics for the sample. The mean parcel size for a household is about 0.15 Ha, with a maximum size of less than 3.7 Ha. Harvest labor is disproportionately skewed to family participation with around 20 days of harvest labor being conducted by family members and only about 2 days for hired labor. This is indicative of either \textbf{malfunctioning} labor markets, income constraints or high supervision costs. The distances to some key areas are also highlighted in Table \ref{tbl:summary}. On average, households are around 13 kilometers away from an asphalt road and almost 60 kilometers away from the nearest market. This makes both the purchase of improved seed and the marketing of yields subject to transportation costs and a difficult enterprise. Fertilizer costs are on average 8,350 Birr ($\thicksim$178 USD), with a large amount of the sample not using fertilizer at all. This is particularly important in the context of improved maize seed use as these technologies are intensive in external inputs, such as chemical fertilizers and irrigation, to be effective. Only 5\% of the sample irrigates their crops making households susceptible to weather shocks.

@@

A seed is considered “improved” in Ethiopia if it was tested by breeders and evaluated to be superior to existing varieties  (\citealp{MoA13}; \citealp{kosmowski2020shining}). In Ethiopia, the government controls most of the seed system, with local agricultural offices or extension agents assessing the demands of different varieties. The MoA decides the seed quantities to be produced, and actors such as regional seed enterprises, seed companies, seed unions, are involved in the production of certified seeds, which are later distributed by seed producer cooperatives (SPCs).\footnote{They account for 37\% of the total distribution in the country (\citealp{kosmowski2020shining}).} An improved seed can be obtained directly from such distribution centers (which would distribute a first generation seed), however it is often the case that seeds cultivated from second or later generations from the originally bred seed are also exchanged between farmers. In the survey, farmers are asked to report whether the variety they used was a first-generation improved seed, second generation, recycled, or a traditional variety. When buying seeds, farmers might be able to identify the variety they are looking for, but it is unlikely that they would be aware of the source.

Fifty four varieties have been released in Ethiopia since 1990.\footnote{Out of which seven were released between 1990 to 1999, 29 between 2000 to 2009, and 18 between 2010 to 2019.} Thirty four such varieties contain International Maize and Wheat Improvement Center (CIMMYT) germplasm, out of which, 10 Drought-tolerant maize varieties (DTMZ) and 8 Quality protein maize (QPM) were released (most of these between 2010 to 2019). Besides the CIMMYT-related germplasm, ten are open-pollinated varieties (OPVs) and 25 are hybrid (\citealp{kosmowski2020shining}).

The fourth wave of the ESS introduced DNA fingerprinting of barley, maize, and sorghum for the first time. The collection of cropcuts for crop yield estimation facilitated the process for DNA fingerprinting. For the cropcuts, a four square meter quadrant was randomly selected in each plot, and total production harvested from the quadrant was weighed (both when freshly harvested and when dried). The DNA material was extracted from samples from the dried cropcuts, and were then matched with the genetic material of the maize DNA reference library for Ethiopia, obtained from a CIMMYT and Ethiopian Institute of Agricultural Research (EIAR) project. The collection of DNA samples focused on eight regions. In each enumeration area (EA) of these regions, a random sample of plots was selected, with a maximum of 10 fields for each crop. As a result, 505 plots were selected for DNA fingerprinting, corresponding to 447 households.

@@

We find that 38\% of households self-report adopting a “new” improved variety (SR2 in Figure \ref{fig:adoption_r4}), and 43\% report adopting a “new”, “second generation” or “recycled” improved variety (SR1). Considering a household as adopter if the household planted an improved variety in at least one of their plots, leads to classifying all households where a DNA sample was obtained as users of improved seed variety, a result that holds when the purity threshold of 70\% (DNA 70) is used. It is only when using purity thresholds of 90 and 95\% (DNA 90, DNA 95 in Figure \ref{fig:adoption_r4}, respectively), that adoption rates of improved varieties change to 91 and 54\% of sampled households, respectively.\footnote{We also find that 62\% of households adopted a hybrid variety, compared to 38\% that adopted an open-pollinated variety, and 15\% of households adopted a drought-tolerant maize variety, while 63\% of households that adopted an improved variety use a CGIAR-derived germplasm, and 79\% use an exotic germplasm.} Finally, 64\% of households adopted a variety that was released in the year 2000 or after, and 24\% adopted a variety that was adopted in the year 2010 or after. These ratios are in line with prior findings (\citealp{Zeng15}).

@@

We find regional variations in adoption levels. In the regions of Oromia, Amhara, Tigray, and SNNPR, around 39\% of households report adopting an improved varierty . In the regions of Amhara, Benishangul-Gumuz, Oromia, SNNPR, and Tigray, adoption rates of improved maize varieties were found for 27\% of households (\citealp{Jaleta2018-oj}). In Oromia, SNNPR, and Benishangul-Gumuz adoption of improved varieties per farmers’ self-reports was found for 84\% of maize growers in 2011, and 88\% of maize growers in 2013 (\citealp{Yirga17}).

@@

We use self-reported and cropcut yields as outcomes in a Group Random Coefficients (GRC) approach to investigate the heterogeneous effects of adoption. The advantage of the GRC approach is that we do not rely on identification from a valid instrument, but on modeling the heterogeneity directly through a projection of adoption returns on their observed ``trajectory'' of adoption. This allows us to directly capture two effects: heterogeneous effects in the form of the so-called ``comparative advantage" of adoption, and, under the assumption of linearity in comparative advantage, a description of the overall economy's sorting of farmers based on their comparative advantage. The first is important because it is able to disentangle the effect of a general household's agricultural ability (which can be thought of as a fixed effect) and their adoption-specific potential. The second parameter can tell us whether the economy sort farmers in a way that is commensurate with providing those that would benefit the most from adoption actually adopting, or whether there are barriers interfering with that. See \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} for more information on the assumptions involved for the GRC approach.

@@

The endogenous switching regression holds certain advantages over other methodologies used for identifying unbiased estimates with cross-sectional observational data such as propensity score matching or other instrumental variable approaches (\citealp{Shiferaw2014-op}) and has been used to study similar questions in the past (\citealp{falco2011does}; \citealp{kabunga2012yield}). \par
This approach models the effect of improved seed adoption on yields in two stages. First, the decision to adopt an improved seed is determined by a series of observed and unobserved variables. Equation \ref{eq:switch_latent} shows the definition for a latent variable $A^*$ that reflects the benefit from adopting an improved seed relative to planting a traditional one. As shown in the equation, household $i$ will decide to adopt ($A_i$=1) if these perceived benefits of adoption are positive.

@@

The results of the GRC regression are shown in Table \ref{tbl:unres}. The results for self-reported and cropcut yields are largely similar. There is some discrepancy in the self-reported yields in the first wave, and specifically with the $(1,0,0)$ trajectory. The return to adoption is negative for cropcuts, but positive for self-reported yields. The returns to adoption in many cases, are not statistically significant, or they are negative. Only in the case of the trajectory $(0,0,1)$: those that adopted in the second and third waves, is there a large, statistically significant effect.
